The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Converted to PASA format

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{xspace}
\usepackage{subcaption}

\usepackage{fixltx2e} % fixes placing of image positions

@@

\ifpdf\pdfinfo{/Title      (Analysis of the Murchison radio environment for low-frequency astronomy)
               /Author     (A. R. Offringa et al.)
               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

@@

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~Kaplan\Curtin,
... , %\newauthor
builders list
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
}

\begin{document}

\date{Accepted TODO. Received TODO; in original form TODO}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pubyear{2014}

\label{firstpage}
\maketitle

@@

The \textsc{aoflagger} software provides a C++ library\footnote{\url{http://aoflagger.sourceforge.net/doc/api/}} that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those. }\label{tbl:obs-list}
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & ---\\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & -- \\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 3\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & ---\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{...\%} & \textbf{0.5\%}\\
\hline
\end{tabular}
\end{table*}

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.

The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.

@@

\section{Detection results}

@@

\bibliographystyle{mn2e}
\bibliography{references}

>>>>

\documentclass{pasa}
\bibliographystyle{apj}
%\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
%\usepackage{xspace}
\usepackage{subcaption}

%\usepackage{fixltx2e} % fixes placing of image positions

@@

\ifpdf\pdfinfo{/Title      (Analysis of the Murchison radio environment for low-frequency astronomy)
               /Author     (A. R. Offringa et al.)
               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
        }
\else\usepackage{graphics}\fi
\usepackage{graphics}

@@

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~Kaplan...,
... , %\newauthor
builders list
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

\begin{document}
\maketitle

%\date{Accepted TODO. Received TODO; in original form TODO}
%\pagerange{\pageref{firstpage}--\pageref{lastpage}}
%\pubyear{2014}

\label{firstpage}

@@

The \textsc{aoflagger} software provides a C++ library%TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

\begin{table*}
\caption{List of observations used in the analyses. The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives). The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag. Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those. }\label{tbl:obs-list}
\begin{tabular}{|l|l|r|r|r|c|}
\hline
\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
\hline
EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & ---\\
GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & -- \\
GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 3\% \\
EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & ---\\
GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
\hline
\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.5\%}\\
\hline
\end{tabular}
\end{table*}

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.

@@

The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature. This shape is e.g. visible in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands. This causes a direction-dependent band-pass. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the number of removed edge channels to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

\section{Detection results}
The second last column of Table~\ref{tbl:obs-list} list for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}.

@@

\bibliography{references}
